### 2020-05-31 paper.tex
Edit Internal Review

Debuggers available in IDEs and the developer tools of a web browser\footnote{\url{https://developer.mozilla.org/en-US/docs/Tools/Debugger}} today are specialized to work with imperative, control-flow oriented programs. E.g., when an engineer wants to know which part of the program called a particular function, the call stack gives a clear answer to this question. Assuming a dataflow oriented program implemented using RP, the question, which transformation was executed right before another cannot be answered using a call stack: A breakpoint within the transformation will not point to its logical predecessor; it will lead to the internals of the RP runtime instead.

This lack of specific knowledge about RP concepts in traditional debugging utilities does often end in time-consuming debugging sessions, as we can confirm from our own experience. Previous research has looked into this field \cite{10.1145/2577080.2577083} \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} over the past years already. We will revisit their results in this paper and elaborate on the particular challenges of debugging RAs implemented using RxJS.

@@

\begin{itemize}
	\item \emph{RQ3: What is the impact of a proposed solution on the debugging experience of a software engineer?}
\end{itemize}

We will conclude this introductory section with the clarification of important terms. Section~\ref{sec:interviews} presents the interview results along with the collected war stories and formalizes two recurrent challenges. With Section~\ref{sec:discussion}, we will have a look at current and previous research regarding the debugging of RAs as well as specific tooling. We are going to validate presented challenges against currently available tools within Section~\ref{sec:study}, show threats to the validity of our work in Section~\ref{sec:threats} and finally end with an outlook on future iterations.

@@

Various ways of implementing RAs emerged over time. The Observer design pattern \cite{gamma1995design}, originated in Object-Oriented Programming \emph{(OOP)}, is a pattern we can find in many of today's API implementations\cite{alabor:2019:reactiveappllications} allowing reactive behavior. Functional Programming \emph{(FP)} shines with specific reactive language extensions like Fran (Haskell) \cite{10.1145/2501654.2501666} or more recently REScala (Scala) \cite{10.1145/2577080.2577083}.

RP is a declarative programming paradigm. Where engineers use imperative programming languages to tell a program every step \emph{how} to do something, a declarative language allows the description of \emph{what} wants to be achieved. The language runtime system then figures out a way to satisfy that given description. RP environments specifically provide abstractions to describe a dataflow and dependency graph used to transform and direct data at runtime.

An RP environment provides (i) language constructs to describe the mentioned graphs, often in the form of a DSL\cite{10.1145/2577080.2577083}. At program runtime, a (ii) runtime system takes care that changed values get propagated through the graph, transformations recomputed where necessary and a consistent application state is maintained.

@@

We divide debugging concepts roughly into three categories: Traditional (i) \emph{imperative-focused debuggers} provide necessary functionality to interact with programs at runtime: Once code execution is paused (i.e. using a breakpoint at a given statement), we gain access to the current call stack and the values assigned to variables of a given frame. Manual control of the program execution allows inspecting its behavior step by step as well as assigning new values to variables ``on-the-fly.''

@@

Traditional, as well as reactive debuggers work with the \emph{current} state of a program's execution only. They lack knowledge about what happened before or what is going to happen in the future. This shortcoming is tedious, especially when debugging a problem depending on many complex circumstances in a system. An (iii) \emph{omniscient debugger} \cite{5287015} \cite{DBLP:journals/corr/OCallahanJFHNP17} does not interact with the executed program directly. Instead, it records runtime telemetry and provides an interface for later inspection. Engineers can ``time travel'' back and forth through the program execution trace without the hassle of reconstructing a given failure situation over and over again. This concept comes with its own set of challenges though: Once decided \emph{what} information should be recorded (everything, only specific parts of the program, one thread, multiple threads, \dots), demands on storage performance and capacity increase while the pool of collected data grows. As a result, user experience (\emph{UX}) concepts need to be aligned for the engineers being able to interact with the sheer amount\cite{5287015} of information efficiently.

@@

``Reactive Extensions'' (\emph{ReactiveX}) is an open-source project. Its members and contributors defined a generic description of an RP API. They further provide reference implementations of this API along with RP runtime environments for various programming languages like Java, C\# and JavaScript\footnote{\url{http://reactivex.io/languages.html}}. ReactiveX summarizes the API as ``\dots a combination of the best ideas from the Observer pattern, the Iterator pattern, and functional programming''\cite{reactivex}. The core concept of the API specification is the Observable\footnote{There is no known relation between ReactiveX' concept of the Observable and the deprecated Java class \href{https://docs.oracle.com/en/java/javase/11/docs/api/java.base/java/util/Observable.html}{\mintinline{TypeScript}{java.util.Observable}}.}: An Observable can be composed and linked (``piped'') with other Observables. It can push (``emit'') values to its dependents. Multiple depending Observables form and describe a dataflow graph eventually. A multitude of operators allows the transformation of values and (higher-order\footnote{An Observable emitted by another Observable is considered a Higher-Order Observable. This naming is related to the concept of higher-order functions in mathematics and computer science.}) Observables. Values emitted from an Observable can be subscribed to at any point, which is closely related to the Observer design patterns \mintinline{Java}{attach} method.

@@

We have worked together with professional software engineers during the process of answering RQ1. On the way of finding our interview partners and war stories reporters, we noticed it to be a challenge on its own to find people who understand themselves as users of RP and related technologies. E.g., even though Angular makes heavy use of RxJS, we will see that many engineers do not directly interact with its abstractions to build simple UIs. In the end, we were able to gather compelling statements which we are going to present in this section.

@@

Four of our interview partners declared to currently work or more recently have worked with RxJS in conjunction with Angular and/or ngrx\footnote{\url{https://ngrx.io/}} to develop frontend web applications. By being the creator of a visualizer for RxJS Observables, the fifth interviewee was a proficient RxJS user. Our sixth interview partner was finally a backend engineer who used akka-streams\footnote{\url{https://akka.io/}} in Scala to model data-flows for a WebSocket-based\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/API/WebSockets_API}}, reactive API layer serving a web frontend application.

@@

\begin{figure}[H]
	\centering
	\input{figures/marble-diagram.pdf_tex}
	\caption{A Marble Diagram visualizing the Observable in Listing~\ref{lst:rxjs-debugging}. Each marble represents an emitted value. The vertical line at marble \mintinline{TypeScript}{6} indicates the Observable completed.}
	\label{fig:marble-diagram}
\end{figure}

@@

Even though the share of non-productive, testing-related code necessary to build mature RxJS-based applications was mentioned to have decreased today, multiple of our reporters mentioned the common practice of manually modifying production code during debugging sessions, hence confirming earlier statements from our informal interview partners. One engineer described a specific scenario where they suspected a problem within a complex Observable composition: Having multiple asynchronous, remote data sources, they used Observables to model the dependencies between them and implemented computations on their results as operators. On top, each data source could re-emit updated versions of previously requested information at any time. Though tested thoroughly, after a week in production, the first of many bugs were reported: ``Figures were changing multiple times (where I expected them to change once), other numbers were not updated at all,'' they told us. The browsers debugger tools and its breakpoints did not help much since the operators were executed several multiple times and some parts of the stream were impossible to grasp, even with conditional breakpoints. ``I started to inspect the flow [\dots] with console.logs and later also using tags from rxjs-spy which exposed more detailed life cycle information'', hence manually augmenting the code with trace logs. After the time-consuming log analysis, they finally were able to resolve the bug. The log statements added were removed in the aftermath, though the rxjs-spy related code was left in the Observable stream in case they might be needed again in the future. (We will elaborate on the functionality of rxjs-spy in Section~\ref{sec:discussion} more in-depth.)

@@

A final story described yet another variation of the application of an external tool: Rather than using a dedicated sandbox to develop pieces of a more complex system, this engineer used Rx Visualizer, an online visualization utilities to generate previously presented marble diagrams from real code. Like in the report before, it was necessary to extract parts from the codebase of the actual application. But once done, the visualizations helped a lot in order to understand when values were emitted, when subscriptions changed and when Observables completed: ``Marble diagrams were a huge help to dig into detailed runtime and life cycle behaviors of the Observables.''.

@@

The sentiment in both the informal interviews and the reported war stories in regards to RP with RxJS was positive. The software engineers value how they can describe dataflows using a DSL, even though the learning curve was perceived as steep. We heard some interesting reports on how RxJS is applied in a daily development environment: Marble diagrams help to understand  an Observables behavior and are useful to implement tests. Large frameworks like Angular hide some of the complexity of RxJS but allow engineers to make use of its full power once the provided functionality needs to be extended.

@@

\begin{listing}[H]
	\begin{minted}{TypeScript}
		interval(1000).pipe(    // Increasing numbers with 1s delay
		  map(i => i * 2),      // Breakpoint possible within i * 2
		  take(4),              // No breakpoint possible
		  tap(console.log)      // Manually added log statement
		).subscribe(showValue); // 0, 2, 4, 6
	\end{minted}
	\caption{Debugging of an RxJS Observable using breakpoints and log statements. \mintinline{TypeScript}{showValue} renders an emitted value to the UI.}
	\label{lst:rxjs-debugging}
\end{listing}

Where a breakpoint can easily be added to Line~2 within the arrow function, this is impossible for \mintinline{TypeScript}{take} on Line~3. One would need to place the breakpoint within the operators' internal implementation instead which can be cumbersome in case the operator is used in a different stream as well. Once the breakpoint on Line~2 interrupts the execution of the program, we will notice another shortcoming related to this circumstance: Rather than representing the logical flow implemented using the DSL, the call stack as shown in Listing~\ref{lst:rxjs-call-stack} points deep into RxJS' internal implementation.

@@

Salvaneschi et al. did focus their research heavily on RP during the mid 2010s. Besides the creation of REScala in 2014 \cite{10.1145/2577080.2577083}, an RP DSL and runtime for the Scala programming language, they have looked into the debugging habits of software engineers as well. In the paper ``Debugging for Reactive Programming'' \cite{10.1145/2884781.2884815}, Salvaneschi et al. elaborate on the necessity of RP debugging tools that understand the specific language abstractions. They provide with the \emph{Reactive Inspector} plugin for the Eclipse IDE a reactive debugger for the REScala RP runtime environment. One of its main features is the visualization of the internal dependency graph. The new class of  breakpoints, the ``reactive breakpoints,'' builds upon the visualized graph as well: They do not depend on a specific line of code or statement rather than on a specific node in the graph. The execution of the program can be interrupted e.g., once a node retains a specific output value or an event of interest takes place in the runtime. Along with the RP specific debugging features, Reactive Inspector implements omniscient debugging concepts as well: Time travel allows to record and replay runtime behavior at a later time without re-execution.

Salvaneschi et al. claim\cite{10.1145/2577080.2577083} the architecture behind \emph{Reactive Inspector} to be flexible enough to be used with different RP runtime environments as well.

@@

\emph{rxfiddle} as proposed by Banken et al. is the first reference implementation of their RP debugger architecture for the ReactiveX API specification described in their paper ``Debugging Data Flows in Reactive Programs'' \cite{10.1145/3180155.3180156}. The design defines two independent parts: The (i) \emph{Host instrumentation} augments a ReactiveX implementation to emit events at runtime (e.g., when an Observable is created or subscribed) and collects them. The (ii) \emph{Visualizer} component interprets these events and presents them along two dimensions: The StoryFlow graph \cite{YWu2013a} shows when an Observable is created and how it interacts with other Observables. A marble diagram visualizes the values emitted over time for every Observable further.

The reference implementation supports event collection for RxJS-based code in versions 4 and 5. It can either be used on a local development machine on existing projects or by copying code snippets to the online available.

The architectural design of \emph{rxfiddle} is similar to the one \emph{Reactive Inspector}: Banken et al. provide a system where different RP runtime environments should be able to be attached to the Visualizer easily.

@@

Software engineers tend to augment their code with additional log trace statements where they have to track down a problem we saw earlier. rxjs-spy\cite{rxjsspy} intends to improve on this by introducing the \mintinline{TypeScript}{tag} operator, along with sophisticated log monitoring facilities. Ideally, ``tagged'' Observables are created during development when the data-flows are composed for the first time. Like the \mintinline{TypeScript}{tap} operator, a tag as shown in Listing~\ref{lst:rxjs-spy-tag} on Line~3 does not influence runtime behavior of an application.

\begin{listing}[H]
	\begin{minted}{TypeScript}
		interval(1000).pipe(
		  map(i => i * 2),
		  tag('multiply'),
		  map(i => i - 1),
		  take(2)
		).subscribe();
	\end{minted}
	\caption{Usage of \emph{rxjs-spy} \mintinline{TypeScript}{tag} Operator on Line~3}
	\label{lst:rxjs-spy-tag}
\end{listing}

Once necessary, rxjs-spy can be advised to log events such as subscribe, emit, error, complete and unsubscribe of matching tags directly to the console without touching any Observable code at all. The additional events logged by rxjs-spy were noted invaluable in the interviews presented earlier as well as in our own experience.

@@

The \emph{Rx Visualizer}\cite{rxviz} is a utility available online in the browser. It is an ``animated playground for Rx Observables''\cite{rxviz} and allows the visualization for snippets of RxJS code using marble diagrams. Diagrams are generated over a given, fixed time and available for download as SVG\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/SVG}} files.

@@

Specialized utilities are aware of RP runtime environments and help software engineers to interpret its state better than before using sole traditional debugger tools. Most of them provide a way to represent the dataflow graph either graphically (Reactive Inspector, rxfiddle) or in textual form (rxjs-spy). They visualize the data flowing through the graph over time in marble diagrams (rxfiddle, Rx Visualizer) or directly in a visual graph representation (Reactive Inspector). Some like Reactive Inspector for REScala provide an ``all in one'' solution with breakpoints, omniscient debugging capabilities, and full IDE integration. In contrast, the RxJS-related tools provide small, differentiated feature sets. Most of them require the engineer to extract code from its origin (rxfiddle, Rx Visualizer) or require manual code preparation or modification (rxjs-spy).

@@

We planned to have a one-hour session for the actual experiment with each subject, followed-up by an unattended quantitative after-action survey. We executed the experiment two consecutive blocks of 25 minutes each. We provided a ZIP file containing the source code for two frontend web applications implemented using TypeScript and RxJS along with a test suite at the start of the experiment. Each of these applications was rigged with two to three bugs, which we asked the subjects to identify and fix using whatever debugging techniques they prefer and commonly use. A block was considered as complete once either the test suite signaled all bugs as resolved or the 25 minutes were exceeded. We asked our subjects to act like in a pair programming situation where they ``think out loud'' their thought process. Though we refrained from answering any question related to the ``where'' a bug has to be expected.

@@

For both problems given, we could observe all subjects to make use of manually added trace log statements either with the existing arrow functions or using the \mintinline{TypeScript}{tap} operator. Neither did any of them use additional external tools where they put parts of the code for further inspection nor did they install additional debugging libraries like rxjs-spy.
After the 25 minutes of the second block exceeded, one subject noted that they would have usually started now to decompose the problem into smaller pieces and observe their behavior more isolated.

@@

Talking to professionals during the informal reviews, evaluating the war stories and the hands-on experiences and results from the observational study, allowed us to identify three main challenges software engineers have to overcome when working with RxJS within an application, hence answering RQ1: (i) They want to \emph{understand how and what data flows} through the graph described by the RP DSL, which helps to identify more straightforward problems most of the time. Though, knowing about the life cycle events of Observables was noted as crucial especially when working with more complex, higher-order compositions of Observables. (ii) Where traditional debugging tools provide support suited for an imperative programming style, they fall short when interacting with RP-based code which would require a \emph{deeper understanding of the RP runtimes internal graph-based model}. This circumstance makes it hard for engineers to retrace dataflows and computations. And finally, (iii) even though there are partial solutions available for the two previously described challenges, they usually require the software engineers either to \emph{manually modify source code or extracting parts of it} to an external utility. These additional efforts cost time and bring their own risks since the production code needs to be touched without adding actual value.

@@

Our overall goal is to support engineers during the process of understanding the inner workings of the RxJS data-flow graph. Existing tools and libraries provide some help, though often struggle by means of the UX: Trace logs with relevant information like life cycle events do improve the debugging process as we saw. Digging through hundreds of log entries containing information of multiple Observables gets tedious quickly, however. Filter and search functionalities help but do not resolve all UX challenges here.

Visualizing the data-flow graph work better to show dependencies within the model compared to textual trace logs. Previous research showed that the sheer amount of nodes in an elaborate graph is hard to visualize effectively \cite{10.1145/3180155.3180156} on one side, on the other not as easy to implement efficiently since the mass of collected data grows quickly. Storing a representation of the graph rarely enough since engineers are interested in the values emitted over time per node as well.

We intend to find a solution that combines the power of textual trace logs with the ease-of-use of graphical visualizations for RxJS data-flow graphs.

@@

RP code needs to be augmented with additional functionality so telemetry data can be collected. Other solutions do this using code weaving\cite{10.1145/2884781.2884815} or similar techniques\cite{10.1145/3180155.3180156} but especially in the area of RxJS often manually. We want to make the manual modification of production source code obsolete. rxfiddle proofed such mechanisms to work with RxJS 4 and 5, hence we intend to bring this feature to version 6, and possibly the upcoming major version 7 as well.

@@

Furthermore, this was the first time we designed an observational study of this kind. Though we carefully peer-reviewed the study design with our advisor, we cannot claim the design and its execution to be flawless. E.g., we would use a commonly known business domain for both problems presented to the subjects, which would allow them to understand the functionality of the application in front of them quicker and more focus on resolving the task given.

@@

We have talked to multiple people in order to understand how software engineers apply RP using RxJS for building applications. We have looked into how they debug their applications and could identify several struggles they have to overcome in this process. We could prove that the engineers inevitably tend to modify their source code once they do not understand how the RP runtime propagates values through its internal data-flow graph by observing professionals in an observational study.

On the way, we elaborated on previous, academic as well as non-academical efforts in the field of reactive debugging. We presented a selection of tools which leverage on their knowledge about the internal workings of RP runtime environments.

>>>>

Debuggers available in IDEs and the developer tools of a web browser\footnote{\url{https://developer.mozilla.org/en-US/docs/Tools/Debugger}} are specialized to work with imperative, control-flow oriented programs. E.g., when an engineer wants to know which part of the program called a particular function, the call stack gives a clear answer to this question. Assuming a dataflow oriented program implemented using RP, the question, which transformation was executed right before another cannot be answered using a call stack: A breakpoint within the transformation will not point to its logical predecessor; it will lead to the internals of the RP runtime instead.

This lack of specific knowledge about RP concepts in traditional debugging utilities does often end in time-consuming debugging sessions, as we can confirm from our own experience. Various researchers have looked into this topic \cite{10.1145/2577080.2577083} \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} over the past years already. We will revisit their results in this paper and elaborate on the particular challenges of debugging RAs implemented using RxJS.

@@

\begin{itemize}
	\item \emph{RQ3: What is the impact of proposed solutions on the debugging experience of software engineers?}
\end{itemize}

We will conclude this introductory section with the clarification of important terms. Section~\ref{sec:interviews} presents the interview results along with the collected war story reports. Before we validate them in an observational study in Section~\ref{sec:study}, we will have a look at current and previous research regarding the debugging of RAs in Section~\ref{sec:discussion}. We will finally answer RQ2 in Section~\ref{sec:future} ``Future Work''.

@@

Various ways of implementing RAs emerged over time. The Observer design pattern \cite{gamma1995design}, originated in Object-Oriented Programming \emph{(OOP)}, is a pattern we can find in many of today's API implementations\cite{alabor:2019:reactiveappllications} allowing reactive behavior. FP shines with reactive language extensions like Fran (Haskell) \cite{10.1145/2501654.2501666} or more recently REScala (Scala) \cite{10.1145/2577080.2577083}.

RP is a declarative programming paradigm. Where engineers use imperative programming languages to tell a program every step \emph{how} to do something, a declarative language allows the description of \emph{what} wants to be achieved. The language runtime system then figures out a way to satisfy that given description. RP runtime environments specifically provide abstractions to describe dataflow and dependency graphs used to transform and direct data at runtime.

An RP language extension provides (i) language abstraction to describe the mentioned graph models, often in the form of a DSL\cite{10.1145/2577080.2577083}. At program runtime, a (ii) runtime environment takes care that changed values get propagated through the graph, transformations recomputed where necessary and a consistent application state is maintained.

@@

We divide debugging concepts roughly into three categories: Traditional (i) \emph{imperative-focused debuggers} provide necessary functionality to interact with programs at runtime: Once a breakpoint pauses the program execution, we gain access to the current call stack and the values assigned to variables of a given frame. Manual control of the program execution allows inspecting its behavior step by step as well as assigning new values to variables ``on-the-fly.''

@@

Traditional, as well as reactive debuggers work with the \emph{current} state of a program's execution only. They lack knowledge about what happened before or what is going to happen in the future. This shortcoming is tedious, especially when debugging a problem depending on many complex circumstances in a system. An (iii) \emph{omniscient debugger} \cite{5287015} \cite{DBLP:journals/corr/OCallahanJFHNP17} does not interact with the executed program directly. Instead, it records runtime telemetry and provides an interface for later inspection. Engineers can ``time travel'' back and forth through the program execution trace without the hassle of reconstructing a given failure situation over and over again. This concept comes with its own set of challenges though: Once decided \emph{what} information should be recorded (everything, only specific parts of the program, one thread, multiple threads, \dots), demands on storage performance and capacity increase while the pool of collected data grows. As a result, user experience (\emph{UX}) concepts need to be aligned so engineers can interact with the sheer amount of information efficiently\cite{5287015}.

@@

``Reactive Extensions'' (\emph{ReactiveX}) is an open-source project. Its members and contributors defined a generic description of an RP API. They further provide reference implementations of this API along with RP language extensions for various programming languages like Java, C\#, and JavaScript\footnote{\url{http://reactivex.io/languages.html}}. ReactiveX summarizes the API as ``\dots a combination of the best ideas from the Observer pattern, the Iterator pattern, and functional programming''\cite{reactivex}. The core concept of the API specification is the Observable\footnote{There is no known relation between ReactiveX' concept of the Observable and the deprecated Java class \href{https://docs.oracle.com/en/java/javase/11/docs/api/java.base/java/util/Observable.html}{\mintinline{TypeScript}{java.util.Observable}}.}: An Observable can be composed and linked (``piped'') with other Observables. It can push (``emit'') values to its dependents. Multiple depending Observables form and describe a dataflow graph eventually. A multitude of operators allows the transformation of values and (higher-order\footnote{An Observable emitted by another Observable is considered a Higher-Order Observable. This naming is related to the concept of higher-order functions in mathematics and computer science.}) Observables. Values emitted from an Observable can be subscribed to at any point, which is closely related to the Observer design patterns \mintinline{Java}{attach} method.

@@

On the way of finding our interview partners and war stories reporters, we noticed it to be a challenge on its own to find people who understand themselves as users of RP and related technologies. E.g., even though Angular makes heavy use of RxJS, we will see that many engineers do not directly interact with its abstractions when building ``basic'' UIs. In the end, we were able to gather compelling statements which we are going to present in this section.

@@

Four of our interview partners declared to currently work or more recently have worked with RxJS in conjunction with Angular and/or ngrx\footnote{\url{https://ngrx.io/}} to develop frontend web applications. By being the creator of a visualizer for RxJS Observables, the fifth interviewee was a proficient RxJS user. Our sixth interview partner was finally a backend engineer who used akka-streams\footnote{\url{https://akka.io/}} in Scala to model dataflows for a WebSocket-based\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/API/WebSockets_API}}, reactive API layer serving a web frontend application.

@@

\begin{figure}[H]
	\centering
	\input{figures/marble-diagram.pdf_tex}
	\caption{A Marble Diagram visualizing the Observable in Listing~\ref{lst:rxjs-debugging}. From left to right, each marble represents an emitted value. The vertical line at the last marble indicates that the Observable completed.}
	\label{fig:marble-diagram}
\end{figure}

@@

Even though the share of non-productive, testing-related code necessary to build mature RxJS-based applications was mentioned to have decreased today, multiple of our reporters commented on the common practice of manually modifying production code during debugging sessions, hence confirming earlier statements from our informal interview partners. One engineer described a specific scenario where they suspected a problem within a complex Observable composition: Having multiple asynchronous, remote data sources, they used Observables to model the dependencies between them and implemented computations on their results as operators. On top, each data source could re-emit updated versions of previously requested information at any time. Though tested thoroughly, after a week in production, the first of many bugs were reported: ``Figures were changing multiple times (where I expected them to change once), other numbers were not updated at all,'' they told us. The browsers debugger tools and its breakpoints did not help much since the operators were executed several multiple times and some parts of the stream were impossible to grasp, even with conditional breakpoints. ``I started to inspect the flow [\dots] with console.logs and later also using tags from rxjs-spy which exposed more detailed life cycle information'', hence manually augmenting the code with trace logs. After the time-consuming log analysis, they finally were able to resolve the bug. The log statements added were removed in the aftermath, though the rxjs-spy related code was left in the Observable stream in case they might be needed again in the future. (We will elaborate on the functionality of rxjs-spy in Section~\ref{sec:discussion} more in-depth.)

@@

A final story described yet another way of utilizing an external tool: Rather than using a dedicated sandbox to develop pieces of a more complex system, this engineer used Rx Visualizer, an online visualization utilities to generate previously presented marble diagrams from real code. Like in the report before, it was necessary to extract parts from the codebase of the actual application. But once done, the visualizations helped a lot in order to understand when values were emitted, when subscriptions changed and when Observables completed: ``Marble diagrams were a huge help to dig into detailed runtime and life cycle behaviors of the Observables.''.

@@

The sentiment in both the informal interviews and the reported war stories in regards to RP with RxJS was positive. The software engineers value how they can describe dataflows using a DSL, even though the learning curve was perceived as steep. We heard some interesting reports on how RxJS is applied in a daily development environment: Marble diagrams help to understand  an Observables behavior and are useful to implement tests. Large frameworks like Angular hide some of the complexity of RxJS but allow engineers to make use of its full power once the pre-provided functionality needs to be extended.

@@

\begin{listing}[H]
	\begin{minted}{TypeScript}
		interval(1000).pipe(    // Increasing numbers with 1s delay
		  map(i => i * 2),      // Breakpoint possible within i * 2
		  take(4),              // No breakpoint possible
		  tap(console.log)      // Manually added log statement
		).subscribe(showValue); // Emits: 0, 2, 4, 6
	\end{minted}
	\caption{Debugging of an RxJS Observable using breakpoints and log statements.}
	\label{lst:rxjs-debugging}
\end{listing}

Where a breakpoint can easily be added to Line~2 within the arrow function, this is impossible for \mintinline{TypeScript}{take} on Line~3. One would need to place the breakpoint within the operators' internal implementation instead which can be cumbersome in case the operator is used in a different stream as well. Once the breakpoint on Line~2 interrupts the execution of the program, we will notice another shortcoming related to this circumstance: Rather than representing the logical flow implemented using the DSL, the call stack as shown in Listing~\ref{lst:rxjs-call-stack} points deep into RxJS' internal implementation. That is why the debuggers step controls cannot operate on the dataflow graph as well; It just does not know this level of abstraction.

@@

Salvaneschi et al. did focus their research heavily on RP during the mid 2010s. Besides the creation of REScala in 2014 \cite{10.1145/2577080.2577083}, an RP DSL and runtime for the Scala programming language, they have looked into the debugging habits of software engineers as well. In the paper ``Debugging for Reactive Programming'' \cite{10.1145/2884781.2884815}, Salvaneschi et al. elaborate on the necessity of RP debugging tools that understand the specific language abstractions. They provide with the \emph{Reactive Inspector} plugin for the Eclipse IDE a reactive debugger for the REScala RP runtime environment. One of its main features is the visualization of the internal dataflow graph. A new class of  breakpoints, the ``reactive breakpoints,'' builds upon the visualized graph as well: They do not depend on a specific line of code or statement rather than on a specific node in the graph. The execution of the program can be interrupted e.g., once a node retains a specific output value or an event of interest takes place in the runtime. Along with the RP specific debugging features, Reactive Inspector implements omniscient debugging concepts as well: Time travel allows to record and replay runtime behavior at a later time without re-execution.

Salvaneschi et al. claim\cite{10.1145/2577080.2577083} the architecture behind \emph{Reactive Inspector} to be flexible enough to be used with other RP runtime environments.

@@

\emph{rxfiddle} as proposed by Banken et al. is the first reference implementation of their RP debugger architecture for the ReactiveX API specification described in their paper ``Debugging Data Flows in Reactive Programs'' \cite{10.1145/3180155.3180156}. The design defines two independent parts: The (i) \emph{Host instrumentation} augments a ReactiveX API implementation to emit events at runtime (e.g., when an Observable is created or subscribed) and collects them. The (ii) \emph{Visualizer} component interprets these events and presents them along two dimensions: The StoryFlow graph \cite{YWu2013a} shows when an Observable is created and how it interacts with other Observables. A marble diagram visualizes the values emitted over time for every Observable further.

The reference implementation supports event collection for RxJS-based code in versions 4 and 5. It is available as an online application where snippets of code can be inspected. A proof-of-concept implementation which works on a local development machine is accessible via the projects Git repository.

The architectural design of \emph{rxfiddle} is similar to the one of \emph{Reactive Inspector}: Banken et al. claim that their Visualizer can be adapted to different RP runtime environments as well.

@@

Software engineers tend to augment their code with additional log trace statements where they have to track down a problem we saw earlier. rxjs-spy\cite{rxjsspy} intends to improve on this by introducing the \mintinline{TypeScript}{tag} operator, along with sophisticated log monitoring facilities. Ideally, ``tagged'' Observables are created during development when the dataflows are composed for the first time. Like the \mintinline{TypeScript}{tap} operator, a tag as shown in Listing~\ref{lst:rxjs-spy-tag} on Line~3 does not influence runtime behavior of an application.

\begin{listing}[H]
	\begin{minted}{TypeScript}
		import { create } from 'rxjs-spy';
		import { tag } from 'rxjs-spy/operators';

		const spy = create();  // Create monitor
		spy.log(/multiply/);   // Log tags matching RegEx /multiply/

		interval(1000).pipe(
		  map(i => i * 2),
		  tag('multiply'),     // Tag Observable with "multiply"
		  map(i => i - 1),
		  take(2)
		).subscribe();
	\end{minted}
	\caption{Usage of \emph{rxjs-spy} \mintinline{TypeScript}{tag} Operator on Line~3}
	\label{lst:rxjs-spy-tag}
\end{listing}

Once necessary, rxjs-spy can be advised to log events such as subscribe, emit, error, complete and unsubscribe of matching tags directly to the console (see Listing~\ref{lst:rxjs-spy-log}) without touching any Observable code at all. The additional events logged by rxjs-spy were noted invaluable in the interviews presented earlier as well as in our own experience.

@@

The \emph{Rx Visualizer}\cite{rxviz} is a utility available online in the browser. It is an ``animated playground for Rx Observables''\cite{rxviz} and allows to visualize snippets of RxJS code using marble diagrams. Diagrams are generated over a given, fixed time and available for download as SVG\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/SVG}} files.

@@

Specialized utilities are aware of RP runtime environments and help software engineers to interpret its state better than before using sole traditional debugger tools. Most of them provide a way to represent the dataflow graph either graphically (Reactive Inspector, rxfiddle) or in textual form (rxjs-spy). They visualize the data flowing through the graph over time in marble diagrams (rxfiddle, Rx Visualizer) or directly in a visual graph representation (Reactive Inspector). Whereas Reactive Inspector for REScala provides an ``all-in-one'' solution with reactive breakpoints, omniscient debugging capabilities, and full IDE integration, the RxJS-related tools provide smaller, differentiated feature sets in contrast. Most of them require the engineer to extract code from its origin (rxfiddle, Rx Visualizer) or call for manual code preparation or modification (rxjs-spy).

@@

We planned to have a one-hour session for the actual experiment with each subject, followed-up by an unattended quantitative after-action survey. We executed the experiment in two consecutive blocks of 25 minutes each. We provided a ZIP file containing the source code for two frontend web applications implemented using TypeScript and RxJS along with a test suite at the start of a session. Each of these applications was rigged with two to three bugs, which we asked the subjects to identify and fix using whatever debugging techniques they prefer and commonly use. A block was considered as complete once either the test suite signaled all bugs as resolved or the 25 minutes were exceeded. We asked our subjects to act like in a pair programming situation where they ``think out loud'' their thought process. Though we refrained from answering any question related to the ``where'' a bug has to be expected.

@@

For both problems given, we could observe all subjects to make use of manually added trace log statements either within an existing arrow function or using the \mintinline{TypeScript}{tap} operator. Neither did any of them use additional external tools where they put parts of the code for further inspection nor did they install additional debugging libraries like rxjs-spy.
After the 25 minutes of the second block exceeded, one subject noted that they would have usually started now to decompose the problem into smaller pieces and observe their behavior more isolated.

@@

Interviewing professionals, evaluating the war stories, the hands-on experiences, and the results from the observational study allowed us to identify three main challenges, software engineers have to overcome when working with RxJS within an application: (i) They want to \emph{understand how and what data flows} through the graph described by the RP DSL, which helps to identify more straightforward problems most of the time. Though, knowing about the life cycle events of Observables was noted as crucial especially when working with more complex, higher-order compositions of Observables. (ii) Where traditional debugging tools provide support suited for an imperative programming style, they fall short when interacting with RP-based code which would require a \emph{deeper understanding of the RP runtimes internal graph-based model}. This circumstance makes it hard for engineers to retrace dataflows and computations. And finally, (iii) even though there are partial solutions available for the two previously described challenges, they usually require the software engineers either to \emph{manually modify source code or extracting parts of it} to an external utility. These additional efforts cost time and bring their own risks since the production code needs to be touched without adding actual value.

@@

Our overall goal is to support engineers during the process of understanding the inner workings of the RxJS dataflow graph. Existing tools and libraries provide some help, though often struggle by means of the UX: Trace logs with relevant information like life cycle events do improve the debugging process as we saw. Digging through hundreds of log entries containing information of multiple Observables gets tedious quickly, however. Filter and search functionalities help but cannot resolve all UX challenges.

A graphical representation of the dataflow graph is better suited to show dependencies within the model then textual trace logs. Previous research showed that the sheer amount of nodes in an elaborate graph is hard to visualize effectively \cite{10.1145/3180155.3180156} on one side, on the other also not easily implement efficiently as well.

We intend to find a solution that combines the power of textual trace logs with the ease-of-use of graphical visualizations for RxJS dataflow graphs. We want this solution to handle small as well as large graphs.

@@

RP code needs to be augmented with additional functionality so telemetry data can be collected. Previous solutions do this using code weaving\cite{10.1145/2884781.2884815} or similar techniques\cite{10.1145/3180155.3180156}. For engineers themselves though, this is often a manual process. We want to make the manual modification of production source code obsolete. rxfiddle proofed such mechanisms to work with RxJS 4 and 5, hence we intend to bring this feature to version 6, and possibly the upcoming major version 7 as well.

@@

Furthermore, this was the first time we designed an observational study of this kind. Though we carefully peer-reviewed the study design with our advisor, we cannot claim the design and its execution to be flawless. E.g., we would use a commonly known business domain for both problems presented to the subjects. This would allow them to understand the functionality of the application in front of them quicker and focus more on resolving the task given.

@@

We have talked to multiple people in order to understand how software engineers apply RP using RxJS for building applications. We have looked into how they debug their applications and could identify several struggles they have to overcome in this process. We could prove that the engineers inevitably tend to modify their source code once they do not understand how the RP runtime propagates values through its internal dataflow graph by observing professionals in an observational study.

On the way, we elaborated on previous, academic as well as non-academic efforts in the field of reactive debugging. We presented a selection of tools which leverage on their knowledge about the internal workings of RP runtime environments.
